Revisiting the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation using complex Empirical Orthogonal Functions

### 2022-03-23 analysis/paper/shceof.Rmd
Usa visual editor.

<!-- The large-scale extratropical circulation in the Southern Hemisphere (SH) is strongly zonally symmetric, but its zonal departures are highly relevant for regional impacts and could even be related to the occurrence of high-impact climate extremes (REFERENCE).  -->
<!-- Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum [@trenberth1980]. -->

@@

<!-- Comparan eventos ENSO del 97 y de 2002 -->
<!-- La posición de las anomalías de circulación (PSA) son importantes para determinar el impacto del ENSO en sudamérica. Sugiere que la localización de la anomalía de SST cambia la localización del PSA.  -->

@@

<!-- > However, the spatial pattern of the PSA wave train excited by ENSO seems to be -->
<!-- > non-stationary. Fogt and Bromwich (2006) indicated that the PSA mode moved -->
<!-- > eastwards by 20° longitude in the 1990s owing to the eastward shift of the tropical -->
<!-- > convective position. In addition, Wilson et al. (2014, 2016) and Ciasto et al. (2015) -->
<!-- > found that the different types of ENSO can contribute to the wave shift. Specifically, -->
<!-- > Rossby waves excited by central Pacific (CP)–type El Niño shift westwards owing to -->
<!-- > the westward movement of the tropical diabatic heating. -->

@@

There is no significant simultaneous correlation between cEOFs time series.
Both cEOFs show year-to-year variability but show no evidence of decadal variability (Fig. \@ref(fig:extended-series)).
Because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean, the cEOFs have non zero temporal mean. 
However, cEOF2’s mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.

@@

(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @@campitelli2022 for the 1979 -- 2019 period. Thick lines represent estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].

@@

The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh) and their comparison with those obtained for SAM, A-SAM and S-SAM by @@campitelli2022.

@@

(ref:enso-phase-cap) ONI plotted against cEOF2 phase for th 1979 -- 2019 period.
Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange diamonds and green circles and respectively. 
Black line is the fit ONI ~ sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.

@@

<!-- To better understand the relationship between precipitation anomalies over eastern Australia and the cEOF2, Figure \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over eastern Australia (box in Figure \@ref(fig:pp-oceania)) as a function of cEOF2 phase. -->
<!-- The black line shows a sinusoidal fit, whose coefficients are fitted by weighted least squares using the magnitude of the EFO2 as weight. -->
<!-- The $r^2$ corresponding to the fit is `r signif(summary(model)$r.squared, 2)` (p-value `r report_p(p_value)`). -->
<!-- There is a strong relationship between the phase of the cEOF2 and precipitation anomalies in eastern Australia, but the Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship. -->

Over South America, the Imaginary cEOF has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil. This again is a well-known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@@campitelli2022].

@@

There is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies [@irving2016].
We show not only that the cEOF2 tends to be in the imaginary phase (~PSA1) when the ENSO region is warm, but that central Pacific SST anomalies tend to move the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to move it towards positive real phase. 
When ENSO is neutral, the cEOF2 is still active, but with no particular preferred phase.
This mirrors the results of @cai2002, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but the variability of one of the PSA modes was enhanced when adding the ENSO signal.
The magnitude of the cEOF2 appears not to be related to the magnitude of ENSO, except for the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.

@@

* The Oceanic Niño Index is available via NOAA's Climate Prediction Center: <https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>.

* The Dipole Mode Index is available via Global Climate Observing System Working Group on Surface Pressure: <https://psl.noaa.gov/gcos_wgsp/Timeseries/Data/dmi.had.long.data>

@@

(ref:pp-temp-cap-1) Same as Figure \@ref(fig:pp-temp-2) but for cEOF1.
(ref:pp-temp-cap-1) Same as Figure \@ref(fig:pp-temp-2) but for cEOF1.

>>>>

<!-- The large-scale extratropical circulation in the Southern Hemisphere (SH) is strongly zonally symmetric, but its zonal departures are highly relevant for regional impacts and could even be related to the occurrence of high-impact climate extremes (REFERENCE).  -->

<!-- Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum [@trenberth1980]. -->

@@

<!-- Comparan eventos ENSO del 97 y de 2002 -->

<!-- La posición de las anomalías de circulación (PSA) son importantes para determinar el impacto del ENSO en sudamérica. Sugiere que la localización de la anomalía de SST cambia la localización del PSA.  -->

@@

<!-- > However, the spatial pattern of the PSA wave train excited by ENSO seems to be -->

<!-- > non-stationary. Fogt and Bromwich (2006) indicated that the PSA mode moved -->

<!-- > eastwards by 20° longitude in the 1990s owing to the eastward shift of the tropical -->

<!-- > convective position. In addition, Wilson et al. (2014, 2016) and Ciasto et al. (2015) -->

<!-- > found that the different types of ENSO can contribute to the wave shift. Specifically, -->

<!-- > Rossby waves excited by central Pacific (CP)–type El Niño shift westwards owing to -->

<!-- > the westward movement of the tropical diabatic heating. -->

@@

There is no significant simultaneous correlation between cEOFs time series.
Both cEOFs show year-to-year variability but show no evidence of decadal variability (Fig. \@ref(fig:extended-series)).
Because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean, the cEOFs have non zero temporal mean.
However, cEOF2's mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.

@@

(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \@@campitelli2022 for the 1979 -- 2019 period. Thick lines represent estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].

@@

The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh) and their comparison with those obtained for SAM, A-SAM and S-SAM by \@@campitelli2022.

@@

(ref:enso-phase-cap) ONI plotted against cEOF2 phase for th 1979 -- 2019 period. Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange diamonds and green circles and respectively. Black line is the fit ONI \~ sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.

@@

<!-- To better understand the relationship between precipitation anomalies over eastern Australia and the cEOF2, Figure @ref(fig:australia-pp-phase) plots precipitation anomalies averaged over eastern Australia (box in Figure @ref(fig:pp-oceania)) as a function of cEOF2 phase. -->

<!-- The black line shows a sinusoidal fit, whose coefficients are fitted by weighted least squares using the magnitude of the EFO2 as weight. -->

<!-- The $r^2$ corresponding to the fit is `r signif(summary(model)$r.squared, 2)` (p-value `r report_p(p_value)`). -->

<!-- There is a strong relationship between the phase of the cEOF2 and precipitation anomalies in eastern Australia, but the Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship. -->

Over South America, the Imaginary cEOF has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil.
This again is a well-known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [\@ @campitelli2022].

@@

There is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies [@irving2016].
We show not only that the cEOF2 tends to be in the imaginary phase (\~PSA1) when the ENSO region is warm, but that central Pacific SST anomalies tend to move the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to move it towards positive real phase.
When ENSO is neutral, the cEOF2 is still active, but with no particular preferred phase.
This mirrors the results of @cai2002, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but the variability of one of the PSA modes was enhanced when adding the ENSO signal.
The magnitude of the cEOF2 appears not to be related to the magnitude of ENSO, except for the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.

@@

-   The Oceanic Niño Index is available via NOAA's Climate Prediction Center: <https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>.

-   The Dipole Mode Index is available via Global Climate Observing System Working Group on Surface Pressure: <https://psl.noaa.gov/gcos_wgsp/Timeseries/Data/dmi.had.long.data>

@@

(ref:pp-temp-cap-1) Same as Figure \@ref(fig:pp-temp-2) but for cEOF1.

(ref:pp-temp-cap-1) Same as Figure \@ref(fig:pp-temp-2) but for cEOF1.
